Exploring Compiler Optimization Techniques

### 2014-11-13 doc/paper.tex
Added block pruning optimization and regalloc start

  \section{SSA conversion optimization}
  \subsection{Problem}
Imperative languages usually allow variables and memory to be rewritten. SSA 
inherently, does not allow temporary variables to change their value over time. 
SSA supports several ways to support this model. One is using the \verb+alloca+
instruction to allocate memory and use the \verb+load+/\verb+store+ system,
another, is cleverly using SSA and its $\phi$ nodes. The first translation uses
rewritable memory and is therefore not a strict SSA representation of control
flow.\footnote{The intermediate form used is still SSA compliant, however, even though it
allows rewriting of \textit{memory}. The temporary values (t1 etc.) are
still single-assignment. Allowing rewritable memory is a necessary evil for
imperative languages.}

@@

This algorithm might look quite involved, but its simplicity is actually 
staggering. It simply depends on phiability optimisation. Phiability 
optimisation fills in constants for variables automatically. Consider the first 
fragment's IR before phiability optimisation:

@@

The variables still exist in their crude memory form. However, their values are 
propagated automatically once phiability optimisation occurs:

@@

\subsection{Considerations}
\subsubsection{Platform incompatibilities}
There is a way compiler-based constant folding might stand in the way of the 
programmer. Mostly the compiler can do this when folding
away instructions  operating on floating point operands, because different targets may compute 
floating point operations differently. Therefore cross-compilation becomes an 
issue; if a floating point instruction for target $Y$ normally yielding $V_y$, it 
yields $V_x$ when folded away by target $X$, causing different semantics before and 
after optimisation.\cite{cfld_cross}

A solution to this problem is to implement a floating point virtual machine for 
several targets, that use non IEEE floating point. Targets using IEEE floating 
point can use C99's internal way of computing IEEE floating point operations.
Since implementing such a system is non-trivial, code duplication needs to be 
avoided. If any other optimisation would need to be able to calculate an 
operation on two constants, it should run the same code. Therefore, the actual 
folding computations are performed outside of the optimiser, by a separate 
folding system.

@@

@@

  \bibitem{cfld_cross}
  \emph{Constant folding and cross compilation}. (s.d.) Consulted on 2014/11/11,
  \url{http://en.wikipedia.org/wiki/Constant_folding#Constant_folding_and_cross_compilation}
\end{thebibliography}

>>>>

  \section{SSA conversion optimization}
  \label{sec:ssa}
  \subsection{Problem}
Imperative languages usually allow variables and memory to be rewritten. SSA 
inherently, does not allow temporary variables to change their value over time. 
SSA supports several ways to support this model. One is using the \verb+alloca+
instruction to allocate memory and use the \verb+load+/\verb+store+ system,
another, is cleverly using SSA and its $\phi$ nodes. The first translation uses
rewritable memory and is therefore not a strict SSA representation of control
flow.\footnote{The intermediate form used is still SSA compliant, however, even though it
allows rewriting of \textit{memory}. The temporary values (t1 etc.) are
still single-assignment. Allowing rewritable memory is a necessary evil for
imperative languages.}

@@

This algorithm might look quite involved, but its simplicity is actually 
staggering. It simply depends on SSA conversion optimization (as described in
section~\ref{sec:ssa}). SSA conversion 
optimization fills in constants for variables automatically. Consider the first 
fragment's IR before SSA conversion optimization:

@@

The variables still exist in their crude memory form. However, their values are 
propagated automatically once SSA conversion optimization occurs:

@@

\subsection{Considerations}
\subsubsection{Platform incompatibilities}
There is a way compiler-based constant folding might stand in the way of the 
programmer. Mostly the compiler can do this when folding
away instructions  operating on floating point operands, because different targets may compute 
floating point operations differently. Therefore cross-compilation becomes an 
issue; if a floating point instruction for target $Y$ normally yielding $V_y$, it 
yields $V_x$ when folded away by target $X$, causing different semantics before and 
after optimization.\cite{cfld_cross}

A solution to this problem is to implement a floating point virtual machine for 
several targets, that use non IEEE floating point. Targets using IEEE floating 
point can use C99's internal way of computing IEEE floating point operations.
Since implementing such a system is non-trivial, code duplication needs to be 
avoided. If any other optimization would need to be able to calculate an 
operation on two constants, it should run the same code. Therefore, the actual 
folding computations are performed outside of the optimiser, by a separate 
folding system.

@@

\section{Block pruning}
\subsection{Problem}
When the intermediate form is generated by the front-end, it may leave blocks in
the IR without any predecessors. This is often the result of a language that
allows the programmer to write unreachable code. I.e. the following C:

\begin{lstlisting}
return 0;
return 1;
\end{lstlisting}

There's no was the second line could possibly be reached, and this is even obvious
in the intermediate form. This would be a possible naive translation:

\begin{lstlisting}
L0:	ret((int)0);
L1:	ret((int)1);
\end{lstlisting}

The \verb+L1+ block can be left out in its entirety, improving code size.

\subsection[Implementation]{Implementation\footnote{A reference implementation
can be found in acc as the o\_prune() pass} }
A block can be left out if it has no predecessors and isn't \verb+L0+.
This is a fairly trivial operation. The only consideration might be a $\phi$
node in one of the block's successors. If such a successor has a $\phi$ node with
an item for the block marked for removal, that item must be removed. That's
always the case, however, when removing a block.

  \section{Register allocation}
  \subsection{As an optimization}
It's hard to categorize register allocation as either an optimization or a
general method used in assembly generation. For the purposes of this article
it'll be classified as the former, because it is a way to speed up generated
assembly.

\subsection{Problem}
The original i386 processors had only very few general purpose registers
(\verb+eax+, \verb+ebx+, \verb+ecx+, \verb+edx+, \verb+esi+ and
\verb+edi+).\cite[section~3.4.1]{intel}

@@

  \bibitem{cfld_cross}
  \emph{Constant folding and cross compilation}. (s.d.) Consulted on 2014/11/11,
  \url{http://en.wikipedia.org/wiki/Constant_folding#Constant_folding_and_cross_compilation}

  \bibitem{intel}
  \emph{Intel\textregistered{} 64 and IA-32 Architectures Software Developer's
Manual Volume~1}.
  (2014)

\end{thebibliography}
